\begin{acknowledgements}
This master thesis was written at the University of Oslo, with supervision from Professors Anders Malthe-Sørenssen and Morten Hjort-Jensen, finishing in June 2014.

I initially started my master studies with the goal of doing experimental studies of semiconductors and solar cells, but after attending Morten's course in computational physics, and with some careful nudging from Anders Hafreager, I realized that computational physics was a field that I found much more interesting. After this discovery I quickly changed the direction of my master studies, and ended up at the Computational Physics branch of the Physics department.

I want to thank Anders Malthe-Sørenssen for excellent guidance during my work, and Morten Hjort-Jensen for his supervision. I also want to thank Anders Hafreager, Mathilde Kamperud, Jørgen Trømborg, Jørgen Høgberget, and all others at Computational Physics at the University for many good discussions and talks that made this thesis into what it is.

I would also like to thank the developers of Ovito\cite{stukowski2010ovito}, Inkscape\cite{webinkscape} and Matplotlib for creating such excellent software, for freely distributing the programs, and for making the source code freely available using an open source software license. These programs have been used for creating nearly all graphics and illustrations presented in this thesis.

Lastly, I thank my wonderful wife Silje for keeping me motivated and energized throughout my master studies, and for feeding me with cake.

%``This work was performed on the Abel Cluster, owned by the University of Oslo and the Norwegian metacenter for High Performance Computing (NOTUR), and operated by the Department for Research Computing at USIT, the University of Oslo IT-department. \url{http://www.hpc.uio.no/}'' \todo{most were done on smaug though}


\end{acknowledgements}
