\subsection{Finding a point inside a tetrahedron}
A tetrahedron consists of four points $\bvec a$, $\bvec b$, $\bvec c$, and $\bvec d$, and four faces spanned by the four possible combinations of the four points. For a face spanned by the points $\bvec a$, $\bvec b$, and $\bvec c$ we can find if a point $\bvec P$ is on the same side of the face as the point $\bvec d$ (the point not used to construct the face) by doing some geometry. We first find the normal vector to the surface $\bvec n$ by the cross product 
\begin{align*}
    \bvec n = (\bvec a-\bvec c)(\bvec b-\bvec c).
\end{align*}
We know that the sign of the dot product between this normal vector and another vector going from the plane to a point will give us information about which side of the plane the point is. This means that if two points $\bvec p_1$ and $\bvec p_2$ are on the side of the plane, the dot product between the normal vector and the two vectors
\begin{align*}
    (\bvec p_i - \bvec k),
\end{align*}
where $\bvec k$ is any point in the plane, should have the same sign. So we find the sign of dot products
\begin{align*}
    &\text{sgn}\big(\bvec n\cdot(\bvec P - \bvec k)\big), \\
    &\text{sgn}\big(\bvec n\cdot(\bvec d - \bvec k)\big),
\end{align*}
and if the sign of these dot products is the same, we know that the point $\bvec P$ is on the same side of the face as the point $\bvec d$. We now see that if we do this for all four faces of the tetrahedra, we know that the point $\bvec P$ is inside the tetrahedra if the signs of \emph{all} pairs of dot products are equal.

To implement this for checking which atoms are between two surfaces (with the volume between the surfaces divided into tetrahedra), we express it as a matrix equation. This reduces the calculation of wether a point is inside a tetrahedron to comparing the signs of five matrix determinants.

% Calculating wether a point is inside a tetrahedron can be reduced to comparing the sign of five different matrix determinants. If we have a point $P = (x,y,z)$ and the four vertices of the tetrahedron are $(x_i, y_i, z_i)$ for $i\in \{1,2,3,4\}$, we can find if the point is inside the tetrahedron by checking if the the following five matrix determinants have the same sign
% \begin{align*}
%     &\begin{vmatrix}
%         x_1 & y_1 & z_1 & 1 \\
%         x_2 & y_2 & z_2 & 1 \\
%         x_3 & y_3 & z_3 & 1 \\
%         x_4 & y_4 & z_4 & 1
%     \end{vmatrix},&
%     &\begin{vmatrix}
%         x & y & z & 1 \\
%         x_2 & y_2 & z_2 & 1 \\
%         x_3 & y_3 & z_3 & 1 \\
%         x_4 & y_4 & z_4 & 1
%     \end{vmatrix},&
%     &\begin{vmatrix}
%         x_1 & y_1 & z_1 & 1 \\
%         x & y & z & 1 \\
%         x_3 & y_3 & z_3 & 1 \\
%         x_4 & y_4 & z_4 & 1
%     \end{vmatrix},&
%     \\
% %     \vspace{5mm}
%     \\
%     &\begin{vmatrix}
%         x_1 & y_1 & z_1 & 1 \\
%         x_2 & y_2 & z_2 & 1 \\
%         x & y & z & 1 \\
%         x_4 & y_4 & z_4 & 1
%     \end{vmatrix},&
%     &\begin{vmatrix}
%         x_1 & y_1 & z_1 & 1 \\
%         x_2 & y_2 & z_2 & 1 \\
%         x_3 & y_3 & z_3 & 1 \\
%         x & y & z & 1
%     \end{vmatrix}.&
% \end{align*}
