\chapter{Nosé-Hoover thermostats\label{appendix:nose_hoover}}
\todoao{Clean up Nosé-Hoover thermostat appendix, relate to ensembles chapter}
% the base for a set of \hl{highly advanced} thermostats called Nos\'e-Hoover chains, that sample the canonical ensemble very well, and give very accurate dynamics. \todo{citations} 

We will here derive the Nosé-Hoover thermostat. We closely follow the derivation in section 6.1.2 \cite{frenkel2001understanding}. See also \cite{nose1984unified,hoover1985canonical} for more information.

To show how the thermostat works we need to use the Lagrangian and Hamiltonian formulation of classical mechanics. The Lagrangian $\Lag$ of a classical $N$-body system is defined as the kinetic energy minus the potential energy $U$
\begin{align*}
    \Lag = K - U% = \sum_{i=1}^N \frac{\vec p_i^2}{2m_i} - U(\vec r),
\end{align*}
and what is called the \emph{generalized} momentum $\vec p_i$ of a \emph{generalized} coordinate $\vec q_i$ defined as
\begin{align}
    \vec p_i = \frac{\partial \mathcal{L}}{\partial \dot{\vec q}_i},
    \label{eq:lag_momentum}
\end{align}
where we denote the time derivative by a dot. These generalized coordinates and momenta are not bound to any one coordinate system, and may be any quantitative attribute of the system.

What the Nosé-Hoover thermostat does is to introduce an additional coordinate $s$ to the Lagrangian, creating a virtual, extended system, with the following Lagrangian\cite{nose1984unified}:
\begin{align}
    \mathcal{L}_\text{Nos\'e} = \sum_{i=1}^N \frac{m_i}{2}s^2 \dot{\vec r}_i^2 - U(\vec r) + \frac{Q}{2}\dot s^2 - 3Nk_BT \ln s,
    \label{eq:nose_lag}
\end{align}
where $Q$ is an effective ``mass'' associated with $s$, and $\vec r_i$ is the \emph{generalized} coordinate from earlier, interpreted as a virtual position of an atom using cartesian coordinates. The momenta of this virtual system follow from \cref{eq:nose_lag,eq:lag_momentum} as
\begin{align*}
    &\vec p_i = \frac{\partial \mathcal{L}}{\partial \dot{\vec r}_i} = m_i s^2 \dot{\vec r}_i \\
    &p_s = \frac{\partial \mathcal{L}}{\partial \dot s} = Q\dot s.
\end{align*}
This gives the following Hamiltonian for the extended system
\begin{align}
    \Ham_\text{Nos\'e} = \sum_{i=1}^N \frac{\vec p_i^2}{2m_is^2} + U(\vec r) + \frac{p_s^2}{2Q} + 3Nk_BT \ln s,
    \label{eq:nose_hamiltonian}
\end{align}
It can be shown that we can relate the generalized coordinates to real variables (real variables indicated by a prime) as follows
\begin{align}
    \vec r' &= \vec r \nonumber\\
    \vec p' &= \vec p/s \nonumber\\
    s' &= s \nonumber\\
    \Delta t' &= \Delta t/s. \label{eq:nose_time_scale}
\end{align}
From \cref{eq:nose_time_scale} we see that $s$ can be interpreted as a scaling factor of the time step. \todobo{Something about this scaling? real and virtual measuring times $\tau'$ and $\tau$}

From the Hamiltonian \cref{eq:nose_hamiltonian} we can derive the equations of motion for the virtual variables $\vec r$, $\vec p$, and $t$, and the real variables $\vec r'$, $\vec p'$, and $t'$ \cite{hoover1985canonical}
\begin{align*}
    \dod{\vec r_i'}{t'}     &= s\dod{\vec r_i}{t} = \frac{\vec p_i}{m_is} = \frac{\vec p_i'}{m_i} \\
    \dod{\vec p_i'}{t'}     &= s\dod{\vec (p_i/s)}{t} = \dod{\vec p_i}{t} - \frac{1}{s}\vec p_i\dod{s}{t} = -\frac{\partial U(\vec r')}{\partial \vec r_i'} - \frac{s'p_s'}{Q}\vec p_i' \\
    \frac{1}{s}\dod{s'}{t'} &= \frac{s}{s}\dod{s}{t} = \frac{s'p_s'}{Q} \\
    \dod{(s'p_s'/Q)}{t'}    &= \frac{s}{Q}\dod{p_s}{t} = \left( \sum_{i=1}^N \frac{p_i'^2}{m_i} - 3Nk_BT\right) / Q.
\end{align*}
% \tododo{remove this complicated version? we haven't derived it anyway...}
These equations can further be simplified if we introduce a thermodynamic friction coefficient $\xi = s'p_s'/Q$ and drop the primes. We then get the following equations of motion
\begin{align}
    \xi             &= \frac{sp_s}{Q} \label{eq:nose_xi}\\
    \dot{\vec r}_i  &= \frac{\vec p_i}{m} \label{eq:nose_position}\\
    \dot{\vec p}_i  &= -\dpd{U(\rvec)}{r_i} - \xi \vec p_i \label{eq:nose_momentum}\\
    \dot\xi         &= \left( \sum_{i=1}^N \frac{p_i^2}{m_i} - 3Nk_BT \right) / Q \label{eq:nose_dotxi}%\\
%     \frac{\dot s}{s} &= \dod{\ln s}{t} = \xi
\end{align}
% Note that the last equation is redundant, since eq. 1-4 form a closed set

