\FloatBarrier
\chapter{Results}
\todoao{See \cite{bonnaud2010molecular} for references on density}%
\todoco{In all results that use distance to matrix: refer to \cref{fig:distance_to_matrix_illustration} when comparing to other results.}%
\todoco{Maybe also note that oxygen atoms closer than $\sim 3$~Å are probably bound to the silicon atom, which may come from our passivation, or from natural chemical reactions in the system.}%
Here we present the result of all the measurements we have done. Most of the measurements have been done for 200 different states for each system, with 100 timesteps of 0.050 picoseconds between each state (5 picoseconds between each state). Most measurements have been measured as function of distance to the silica matrix to study the effects of the interactions between water and silia, and we have tried to find the bulk-like behaviour of all measurements for comparison.

We will compare results between the random rough fractures to the results for bulk like-water in reference system \#1 and \#2, and against the two narrow flat reference systems \#3 and \#4, to see if the changes from flat pores to rough fractures have an effect on the water. The results for the rough fracture systems will also be compared against each other to see if the different fracture geometries have any effect.

We will first present and analyze the results from each measurement individually in \crefrange{sec:results_first_sec}{sec:results_last_sec}, and then later in \cref{chap:ending} we will give a summary of the results, and compare the results from the different measurements against each other to see if there are any further conclusions to be drawn.

% \todoa{Find diffusion for one atom?}

% \input{results/density}
% \FloatBarrier
% \input{results/diffusion}
% \FloatBarrier
% \input{results/tetrahedral_order_parameter}
% \FloatBarrier
% \input{results/distance_to_atom}
% % \FloatBarrier
% \input{results/generation_matrix}
% \FloatBarrier

% \section{Area}
% \section{Volume?}

